A family of accessibility measures derived from spatial interaction principles

### 2025-05-17 paper/manuscript.qmd
intro part 1

Accessibility can be seen to represent an evolution of mobility; accessibility may be defined as the "potential of opportunities for [spatial] interaction" [@hansen1959], in contrast to mobility which is simply, movement in itself or 'spatial interaction'. Mobility has been the basis of traditional transportation planning approaches [@ortuzar_2011_modelling]. While accessibility brings a more holistic understanding of combined transportation and land use systems, incorporating both the explicit consideration of destination as opportunity and the concept of _potential_ [@handyMeasuringAccessibilityExploration1997].

The growing interest in accessibility has been accompanied by a boom in scholarly research using different methods and focusing on different research contexts; it has grown to include studies of access to employment [e.g., @karstEvaluationAccessibilityImpacts2003; @grengs2010job; @paez_jobs_2013; @merlin2017competition; @tao_investigating_2020], health care [e.g., @luo2003; @paez_healthcare_2010; @wan2012three; @delamater2013spatial; @boisjoly2017informality; @pereira_2021_geographic], green spaces [@reyesAccessibility2014; @rojas_accessibility_2016; @liang_novel_2024], schools [e.g., @williams_disparities_2014; @romanillosAccessibilitySchoolsSpatial2018; @marques_accessibility_2021], social contacts [e.g., @neutens_human_2007; @farberActivitySpacesMeasurement2012; @farber_2013_social], and regional economic analysis [e.g., @vickermanAccessibility1999; @lopezMeasuring2008; @ribeiro_road_2010; @gutierrez_evaluating_2011] among many other domains of application. In other words, accessibility analysis is used today to broadly understand the potential to reach (or spatially interact with) opportunities that are important to people [@ferreiraReenactingMobilityAccessibility2020]. However, despite its growth in popularity in scholarly works, challenges remain with respect to the more widespread adoption of accessibility in planning practice. For instance, the diversity of accessibility definitions has been flagged by @vanweeAccessible2016, @handy2020, and @kapatsila_resolving_2023. Further, difficulties in the interpretability and communicability of outputs has also been noticed by @geursAccessibilityEvaluationLanduse2004, @vanweeAccessible2016, and @ferreiraReenactingMobilityAccessibility2020.

Furthermore, the adoption of accessibility in planning practice is not necessarily made easier when potential adopters have to contend with a plethora of definitions, each seemingly more sophisticated but less intuitive than the last [@kapatsila_resolving_2023]. At a high level, @geursAccessibilityEvaluationLanduse2004 identify four families of accessibility measures: infrastructure-, place-, person-, and utility-based. Of the place-based family, which is the focus of this paper, the menu has grown to include gravity-based accessibility [e.g., @hansen1959; @pirie_measuring_1979], cumulative opportunities [e.g., @wachs_physical_1973; @pirie_measuring_1979; @ye_spatial_2018], modified gravity [e.g., @SchuurmanMeasuring2010], 2-Step Floating Catchment Areas [e.g., @luo2003], Enhanced 2-Step Floating Catchment Areas [e.g., @luoEnhanced2009], 3-Stage Floating Catchment Areas [e.g., @wan2012three], Modified 2-Step Floating Catchment Areas [e.g., @delamater2013spatial], inverse 2-Step Floating Catchment Areas [e.g., @wang_2sfca_2021], and n-steps Floating Catchment Areas [@liang_novel_2024]. How is a practitioner to choose among this myriad options? What differences in accessibility scores should matter, and how should they be communicated? [see @vanweeAccessible2016; p. 14].

More specifically, this work seeks to address the breath of place-based accessibility's definitions and lack of interpretability by demonstrating that mobility-oriented models--such as the commonly used gravity-based accessibility [e.g., @hansen1959] and the spatial interaction model [e.g., @wilson1971]-- are rooted in the same spatial interaction modeling foundation. In fact, accessibility can be specified using spatial interaction principles and a _family of measures_, akin to the family of spatial interaction models introduced in @wilson1971, can be defined using balancing factors that constrains values based on known information about the land-use.

In this way, this work aims to offer three contributions: (1) introduce and (2) define a family of _constrained_ accessibility measures, namely the total constrained accessibility measure, the singly constrained accessibility measure, and the doubly constrained measure. These measures are explicitly connected to popular measures such as the Hansen-type accessibility [@hansen1959], the popular competition approach of the 2-Step Floating Catchment Area (2SFCA) [@shenLocationCharacteristicsInnercity1998; @luo2003], and the concept of market potential, or what Reilly called 'potential trade territories' [@reilly1929methods] and further developed in other examples [@harris_market_1954; @vickermanAccessibilityAttractionPotential1974]. These Wilson-analogous balancing factors introduced are also defined in order of increasing restrictiveness to shed light on the role of _potential_ in accessibility and access. (3) This work also aims to demonstrate the intuitive value of restoring units to the resulting raw accessibility values. Each zonal and zonal flow value from a constrained accessibility measure is in interpretable units, namely the number of 'opportunities for spatial interaction' or 'population for spatial interaction'. This is in contrast to conventional accessibility measures that yield values in units of 'opportunities weighted by some representation of travel friction'.

@@

Another variant of the total constrained accessibility measure is the _total constrained accessible population measure_, which represents the transpose of $i$ to $j$ of the total constrained accessible opportunities measure. This variant, expressed in @eq-total-constrained-market, represents an expression of the concept of market potential (i.e., potential users) as proposed in @harris_market_1954 and @vickermanAccessibilityAttractionPotential1974. This unconstrained form of market potential $M_j^0$ (@eq-unconstrained-market), effectively the $i$ to $j$ transpose of $V^0_{ij}$, has been used in recent research to express the potentially accessible population (i.e., users) as a result of regional transportation infrastructure investment projects [e.g., @gutierrezLocationEconomicPotential2001; @holl2007twenty; @condecco2018road]. Put another way, market potential can also be thought of as a form of _passive accessibility_, indicating the number of people that can reach each destination. However, like $V_{ij}^0$, issues of unit interpretability arise in $M_j^0$'s unconstrained form. To address this, the constrained variant, the total constrained accessible population measure $M^T_{j}$, is introduced. To formulate this variant, the total balancing factor $K^T$ is applied to the mass of the *population* at $i$ ($W_i^{(1)}$) instead of the opportunities at $j$.

>>>>

Accessibility may be defined as the "potential of opportunities for [spatial] interaction" [@hansen1959], in contrast to mobility which is simply, movement in itself or 'spatial interaction'. Mobility has been the basis of traditional transportation planning approaches [@ortuzar_2011_modelling]. While accessibility brings a more holistic understanding of combined transportation and land use systems, incorporating both the explicit consideration of destination as opportunity and the concept of _potential_ [@handyMeasuringAccessibilityExploration1997].

The growing interest in accessibility has been accompanied by a boom in scholarly research using different methods and focusing on different research contexts; it has grown to include studies of access to employment [e.g., @karstEvaluationAccessibilityImpacts2003; @grengs2010job; @paez_jobs_2013; @merlin2017competition; @tao_investigating_2020], health care [e.g., @luo2003; @paez_healthcare_2010; @wan2012three; @delamater2013spatial; @boisjoly2017informality; @pereira_2021_geographic], green spaces [@reyesAccessibility2014; @rojas_accessibility_2016; @liang_novel_2024], schools [e.g., @williams_disparities_2014; @romanillosAccessibilitySchoolsSpatial2018; @marques_accessibility_2021], social contacts [e.g., @neutens_human_2007; @farberActivitySpacesMeasurement2012; @farber_2013_social], and regional economic analysis [e.g., @vickermanAccessibility1999; @lopezMeasuring2008; @ribeiro_road_2010; @gutierrez_evaluating_2011] among many other domains of application. In other words, accessibility analysis is used today to broadly understand the potential to reach (or spatially interact with) opportunities that are important to people [@ferreiraReenactingMobilityAccessibility2020]. However, despite its growth in popularity in scholarly works, challenges remain with respect to the more widespread adoption of accessibility in planning practice. For instance, the diversity of accessibility definitions has been flagged by @vanweeAccessible2016, @handy2020, and @kapatsila_resolving_2023. Further, difficulties in the interpretability and communicability of outputs has also been noticed by many authores, including @geursAccessibilityEvaluationLanduse2004, @vanweeAccessible2016, and @ferreiraReenactingMobilityAccessibility2020.

The adoption of accessibility in planning practice is not necessarily made easier when potential adopters have to contend with a plethora of definitions, each seemingly more sophisticated but less intuitive than the last [@kapatsila_resolving_2023]. At a high level, @geursAccessibilityEvaluationLanduse2004 identify four families of accessibility measures: infrastructure-, place-, person-, and utility-based. Of the place-based family, which is the focus of this paper, the menu has grown to include gravity-based accessibility [e.g., @hansen1959; @pirie_measuring_1979], cumulative opportunities [e.g., @wachs_physical_1973; @pirie_measuring_1979; @ye_spatial_2018], modified gravity [e.g., @SchuurmanMeasuring2010], 2-Step Floating Catchment Areas [e.g., @luo2003], Enhanced 2-Step Floating Catchment Areas [e.g., @luoEnhanced2009], 3-Stage Floating Catchment Areas [e.g., @wan2012three], Modified 2-Step Floating Catchment Areas [e.g., @delamater2013spatial], inverse 2-Step Floating Catchment Areas [e.g., @wang_2sfca_2021], and n-steps Floating Catchment Areas [@liang_novel_2024]. How is a practitioner to choose among this myriad options? What differences in accessibility scores should matter, and how should they be communicated? [see @vanweeAccessible2016; p. 14].

Here, we seek to address the breath of place-based accessibility's definitions and lack of interpretability by demonstrating that mobility-oriented models--such as the commonly used gravity-based accessibility [e.g., @hansen1959] and the spatial interaction model [e.g., @wilson1971]-- are rooted in the same spatial interaction modeling foundation. In fact, we propose that accessibility can be specified using spatial interaction principles as a _family of measures_, akin to the family of spatial interaction models introduced in @wilson1971 that can be defined using balancing factors that constrains values based on known information about the land-use.

This work aims to offer three contributions: (1) it introduces a family of accessibility measures within the principles of spatial interaction; (2) it formally defines three _constrained_ accessibility measures by reintroducing Wilson-analogous balancing factors. These measures are the total constrained accessibility measure, the singly constrained accessibility measure, and the doubly constrained measure, which are explicitly connected to popular measures such as the Hansen-type accessibility [@hansen1959], the popular competition approach of the 2-Step Floating Catchment Area (2SFCA) [@shenLocationCharacteristicsInnercity1998; @luo2003], and the concept of market potential [@harris_market_1954; @vickermanAccessibilityAttractionPotential1974]. These Wilson-analogous balancing factors introduced are also defined in order of increasing restrictiveness to shed light on the role of _potential_ in accessibility and access. (3) This work also demonstrates that the introduction of balancing factors makes accessibility measures easier to interpret and to communicate by restoring the measurement units to the resulting raw accessibility values. Each zonal and zonal flow value from a constrained accessibility measure is always in interpretable units, namely the number of 'opportunities for spatial interaction' or 'population for spatial interaction'. This is in contrast to conventional accessibility measures, particularly gravity based measures, that yield values in units of 'opportunities weighted by some representation of travel friction'.

@@

Another variant of the total constrained accessibility measure is the _total constrained accessible population measure_, which represents the transpose of $i$ to $j$ of the total constrained accessible opportunities measure. This variant, expressed in @eq-total-constrained-market, represents an expression of the concept of market potential (i.e., potential users) as proposed in @harris_market_1954 and @vickermanAccessibilityAttractionPotential1974, and which Reilly earlier referred to as 'potential trade territories' [@reilly1929methods]. This unconstrained form of market potential $M_j^0$ (@eq-unconstrained-market), effectively the $i$ to $j$ transpose of $V^0_{ij}$, has been used in recent research to express the potentially accessible population (i.e., users) as a result of regional transportation infrastructure investment projects [e.g., @gutierrezLocationEconomicPotential2001; @holl2007twenty; @condecco2018road]. Put another way, market potential can also be thought of as a form of _passive accessibility_, indicating the number of people that can reach each destination. However, like $V_{ij}^0$, issues of unit interpretability arise in $M_j^0$'s unconstrained form. To address this, the constrained variant, the total constrained accessible population measure $M^T_{j}$, is introduced. To formulate this variant, the total balancing factor $K^T$ is applied to the mass of the *population* at $i$ ($W_i^{(1)}$) instead of the opportunities at $j$.
